---
title: "EPPS 6323: Assignment 3 - Regression Plots and Prediction"
author: "JP ADJADEH"
format: html
editor: visual
execute:
  warning: false
  message: false
---

## Load data

```{r}
library(haven)
TEDS_2016 <- read_stata("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")
```

## Create the regplot function

```{r}
regplot = function(x, y) {
  fit = lm(y ~ x)
  plot(x, y)
  abline(fit, col = "red")
}
```

## Regplot: Age vs votetsai

```{r}
regplot(TEDS_2016$age, TEDS_2016$votetsai)
```

## Regplot: Education vs votetsai

```{r}
regplot(TEDS_2016$edu, TEDS_2016$votetsai)
```

## Regplot: Income vs votetsai

```{r}
regplot(TEDS_2016$income, TEDS_2016$votetsai)
```

## What is the problem?

The dependent variable `votetsai` is binary — it only takes values 0 or 1. Linear regression assumes a continuous outcome, so fitting a straight line through binary data doesn't make sense. The regression line can predict values below 0 or above 1, which aren't valid probabilities. With only two categories in the DV, OLS regression isn't the right tool here.

## What can be done to improve prediction?

Since the outcome is binary, logistic regression is the better choice. It models the probability of voting for Tsai as a value between 0 and 1 using a logit link function.

```{r}
model <- glm(votetsai ~ age + edu + income, data = TEDS_2016, family = binomial)
summary(model)
```
