---
title: "Assignment 3: Mapping Census Data"
author: "Your Name"
date: "2025-09-30"
format: 
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-location: left
execute:
  warning: false
  message: false
---

# Census Data Analysis: Texas Counties

## Setup and Data Collection

```{r setup}
#| echo: true
#| output: false

# Load required packages
library(tidycensus)
library(tigris)
library(sf)
library(dplyr)
library(ggplot2)
library(readr)

options(tigris_use_cache = TRUE)

# Set Census API key
census_api_key("b1b5addc32c59ed25b2ac40f0264ab6a93c70fe4", install = FALSE)

# Explore available variables
vars <- load_variables(2023, "acs5", cache = TRUE)
```

```{r parameters}
#| echo: true

# Define parameters
state_abbr <- "TX"
geo_level  <- "county"
my_vars    <- c(income = "B19013_001", poverty = "B17001_002")
year_acs   <- 2023
survey     <- "acs5"
```

```{r download-data}
#| echo: true
#| output: false

# Download ACS data with geometry
acs <- get_acs(
  geography = geo_level,
  variables = my_vars,
  state = state_abbr,
  year = year_acs,
  survey = survey,
  geometry = TRUE
)

# Convert to wide format
acs_wide <- acs |>
  tidyr::pivot_wider(
    id_cols = c(GEOID, NAME, geometry),
    names_from = variable,
    values_from = c(estimate, moe)
  )
```

## Visualization

### Figure 1: Choropleth Map of Median Household Income

```{r choropleth-map}
#| echo: true
#| fig-width: 10
#| fig-height: 8

ggplot(acs_wide) +
  geom_sf(aes(fill = estimate_income), color = NA) +
  scale_fill_viridis_c(name = "Median HH Income") +
  labs(
    title = paste0("ACS ", year_acs, " 5-year: Median Income — ", state_abbr, " (", geo_level, ")"),
    caption = "Source: U.S. Census Bureau via tidycensus"
  ) +
  theme_minimal()
```

**Interpretation:** The choropleth map shows strong geographic variation in median household income across Texas counties. Higher incomes cluster around major metropolitan regions such as Houston, Dallas, Austin, and San Antonio, while many rural areas, especially in West Texas and along the border, report substantially lower incomes. The visualization highlights the spatial concentration of economic advantage in urban centers compared to persistent income disparities in less populated counties.

## Data Tables

### Top 10 Counties by Median Household Income

```{r top-counties}
#| echo: true

top10 <- acs_wide |>
  arrange(desc(estimate_income)) |>
  select(NAME, estimate_income, moe_income) |>
  slice_head(n = 10) |>
  st_drop_geometry()

# Format as table
top10 |>
  mutate(
    Rank = row_number(),
    County = NAME,
    `Estimate Income ($)` = format(estimate_income, big.mark = ","),
    `Margin of Error ($)` = format(moe_income, big.mark = ",")
  ) |>
  select(Rank, County, `Estimate Income ($)`, `Margin of Error ($)`) |>
  knitr::kable()
```

**Note:** Estimates are from the 2023 American Community Survey, 5-year estimates.

The highest-income counties include Harris, Dallas, Bexar, Tarrant, and Travis, reflecting the economic strength of Texas's largest metropolitan areas. These counties benefit from diverse economies, large populations, and employment opportunities that support higher household incomes. The estimates are relatively precise, with moderate margins of error.

### Bottom 10 Counties by Median Household Income

```{r bottom-counties}
#| echo: true

bottom10 <- acs_wide |>
  arrange(estimate_income) |>
  select(NAME, estimate_income, moe_income) |>
  slice_head(n = 10) |>
  st_drop_geometry()

# Format as table
bottom10 |>
  mutate(
    Rank = row_number(),
    County = NAME,
    `Estimate Income ($)` = format(estimate_income, big.mark = ","),
    `Margin of Error ($)` = format(moe_income, big.mark = ",")
  ) |>
  select(Rank, County, `Estimate Income ($)`, `Margin of Error ($)`) |>
  knitr::kable()
```

**Note:** Estimates are from the 2023 American Community Survey, 5-year estimates.

At the other end, counties such as Kenedy, Loving, and King show the lowest median incomes in the state. Many of these counties are sparsely populated, which explains both the very low-income figures and the relatively large margins of error. The data highlight the economic vulnerability of rural Texas, where limited labor markets and small sample sizes constrain reliable measurement.

## Conclusion

Taken together, the map and tables underscore the sharp income inequality across Texas counties. Metropolitan areas consistently report the highest incomes, while rural counties remain at the bottom, with wide confidence intervals that reflect small populations. These results demonstrate how geography and population density shape economic outcomes, pointing to persistent divides between urban prosperity and rural disadvantage.

## Data Export

```{r export}
#| echo: true
#| output: false

# Save the processed data
readr::write_csv(st_drop_geometry(acs_wide), "acs_data.csv")
```

*Note: Processed data saved to `acs_data.csv`*
