---

---
title: "Assignment 5: Data Visualization"
author: "JP ADJADEH"
format:
  html:
    code-fold: true
    code-tools: true
editor: visual
---
## Introduction

This assignment demonstrates data visualization techniques using both base R graphics and ggplot2, applying a professional McKinsey Blue color theme. All visualizations use the built-in `mtcars` dataset.

**Color Palette:**
- McKinsey Blue: #00567D
- McKinsey Light Blue: #0085CA
- McKinsey Gray: #5D5D5D

```{r setup, include=FALSE}
library(ggplot2)

mb <- "#00567D"
ml <- "#0085CA"
mg <- "#5D5D5D"
data(mtcars)
```

---

## Part 1: Base R Graphics

```{r base-setup}
par(col.main = mg, col.lab = mg, cex.main = 1.5, cex.lab = 1.2, las = 1)
```

### 1a. Histogram

```{r histogram-base}
hist(mtcars$mpg, 
     main = "Distribution of MPG", 
     xlab = "MPG", 
     ylab = "Frequency", 
     col = mb, 
     border = "white", 
     breaks = 10)
```

**Interpretation**: The distribution of miles per gallon shows most vehicles cluster around 15-20 MPG, with a right-skewed distribution indicating some high-efficiency outliers.

### 1b(i). Barchart - Vertical

```{r barchart-vertical}
barplot(table(mtcars$cyl), 
        main = "Cars by Cylinder", 
        xlab = "Cylinders", 
        ylab = "Count",
        col = mb, 
        border = "white")
```

**Interpretation**: The dataset contains more 8-cylinder vehicles than any other type, followed by 4-cylinder and 6-cylinder vehicles.

### 1b(ii). Barchart - Horizontal

```{r barchart-horizontal}
barplot(tapply(mtcars$mpg, mtcars$cyl, mean), 
        main = "Avg MPG by Cylinder",
        xlab = "Avg MPG", 
        ylab = "Cylinders", 
        col = ml, 
        border = "white", 
        horiz = TRUE)
```

**Interpretation**: Clear inverse relationship - vehicles with fewer cylinders achieve significantly better fuel efficiency.

### 1c. Pie Chart

```{r piechart}
pie(table(mtcars$cyl), 
    main = "Car Distribution by Cylinder",
    col = c(mb, ml, mg), 
    border = "white")
```

**Interpretation**: 8-cylinder vehicles represent nearly half of the dataset, reflecting the muscle car era when this data was collected (1970s).

### 1d. Boxplot

```{r boxplot-base}
boxplot(hp ~ cyl, 
        data = mtcars, 
        main = "Horsepower by Cylinder",
        xlab = "Cylinders", 
        ylab = "HP", 
        col = mb, 
        border = mg)
```

**Interpretation**: Strong positive correlation between cylinder count and horsepower. 8-cylinder vehicles show the highest median horsepower and greatest variability.

### 1e. Scatterplot

```{r scatterplot-base}
plot(mtcars$wt, mtcars$mpg, 
     main = "Weight vs MPG", 
     xlab = "Weight (1000 lbs)",
     ylab = "MPG", 
     pch = 19, 
     col = mb, 
     cex = 1.5)
abline(lm(mpg ~ wt, data = mtcars), col = ml, lwd = 2)
```

**Interpretation**: Strong negative correlation (r ≈ -0.87) between vehicle weight and fuel efficiency. Each additional 1000 lbs reduces MPG by approximately 5 units.

---

## Part 2: ggplot2 Graphics

```{r ggplot-theme}
theme_mk <- function() {
  theme_minimal() + 
    theme(
      plot.title = element_text(color = mg, size = 16, face = "bold"),
      axis.title = element_text(color = mg, size = 12),
      axis.text = element_text(color = mg),
      panel.grid.major = element_line(color = "lightgray", linetype = "dashed"),
      panel.grid.minor = element_blank()
    )
}
```

### 2a. Histogram

```{r histogram-ggplot}
ggplot(mtcars, aes(mpg)) + 
  geom_histogram(bins = 10, fill = mb, color = "white") +
  labs(title = "Distribution of MPG", x = "MPG", y = "Frequency") + 
  theme_mk()
```

### 2b(i). Barchart - Vertical

```{r barchart-vertical-ggplot}
ggplot(mtcars, aes(factor(cyl))) + 
  geom_bar(fill = mb, color = "white") +
  labs(title = "Cars by Cylinder", x = "Cylinders", y = "Count") + 
  theme_mk()
```

### 2b(ii). Barchart - Horizontal

```{r barchart-horizontal-ggplot}
ggplot(aggregate(mpg ~ cyl, mtcars, mean), aes(factor(cyl), mpg)) +
  geom_bar(stat = "identity", fill = ml, color = "white") + 
  coord_flip() +
  labs(title = "Avg MPG by Cylinder", x = "Cylinders", y = "Avg MPG") + 
  theme_mk()
```

### 2c. Pie Chart

```{r piechart-ggplot}
ggplot(as.data.frame(table(mtcars$cyl)), aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", width = 1, color = "white") + 
  coord_polar("y") +
  scale_fill_manual(values = c(mb, ml, mg)) +
  labs(title = "Car Distribution by Cylinder") + 
  theme_void() +
  theme(plot.title = element_text(color = mg, size = 16, face = "bold", hjust = 0.5))
```

### 2d. Boxplot

```{r boxplot-ggplot}
ggplot(mtcars, aes(factor(cyl), hp)) + 
  geom_boxplot(fill = mb, color = mg) +
  labs(title = "Horsepower by Cylinder", x = "Cylinders", y = "HP") + 
  theme_mk()
```

### 2e. Scatterplot

```{r scatterplot-ggplot}
ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(color = mb, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = ml, size = 1.2) +
  labs(title = "Weight vs MPG", x = "Weight (1000 lbs)", y = "MPG") + 
  theme_mk()
```

---

## Part 3: File Format Exports

Charts can be exported in multiple formats, each with different characteristics and use cases.

```{r export-example, eval=FALSE}
p <- ggplot(mtcars, aes(wt, mpg)) + 
  geom_point(color = mb, size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = ml, size = 1.2) +
  labs(title = "Weight vs MPG", x = "Weight (1000 lbs)", y = "MPG") + 
  theme_mk()

ggsave("scatterplot.pdf", p, width = 8, height = 6)
ggsave("scatterplot.jpg", p, width = 8, height = 6, dpi = 300)
ggsave("scatterplot.svg", p, width = 8, height = 6)
ggsave("scatterplot.tiff", p, width = 8, height = 6, dpi = 300)
ggsave("scatterplot.bmp", p, width = 8, height = 6, dpi = 300)
```

### File Format Comparison

**PDF (Portable Document Format)**
- Type: Vector
- Best for: Print, documents, presentations
- Advantages: Scalable without quality loss, small file size, universal compatibility
- File size: Small (~50KB for typical chart)

**JPG (Joint Photographic Experts Group)**
- Type: Raster (compressed)
- Best for: Web, email, social media
- Advantages: Smallest file size, widely supported
- Disadvantages: Lossy compression, quality degrades with editing
- File size: Smallest (~30KB)

**SVG (Scalable Vector Graphics)**
- Type: Vector
- Best for: Web graphics, interactive visualizations
- Advantages: Scalable, editable in browsers, supports animation
- File size: Small (~60KB)

**TIFF (Tagged Image File Format)**
- Type: Raster (uncompressed or lossless compression)
- Best for: Scientific publishing, archival, high-quality print
- Advantages: Highest quality, lossless, supports layers
- Disadvantages: Very large file size
- File size: Large (~500KB-2MB)

**BMP (Bitmap)**
- Type: Raster (uncompressed)
- Best for: Windows applications, basic image editing
- Advantages: Simple format, no compression artifacts
- Disadvantages: Largest file size, limited features
- File size: Largest (~2-3MB)

**Recommendation**: I will recommend the use of  PDF or SVG for presentations and documents, JPG for web publishing, and TIFF for high-quality print publications.

---

## Conclusion

This assignment successfully demonstrated the creation of five fundamental chart types using both base R graphics and ggplot2, with consistent McKinsey Blue styling applied throughout. 
The professional McKinsey Blue color scheme and consistent styling make these visualizations suitable for business presentations, academic reports, and executive dashboards. The comparison of base R and ggplot2 approaches demonstrates the flexibility and power of R's visualization ecosystem, with ggplot2 offering more refined control over aesthetics while base R provides quick, efficient plotting capabilities.
---
