---
title: "Assignment 1: Data Visualization"
author: "JP ADJADEH"
date: "September 26, 2025"
format: html
---

## 1. Analyzing Anscombe's Plots

Anscombe's Quartet teaches us that data can look the same in numbers but tell very different stories when you make a graph. All four datasets have the same averages, correlations, and regression lines, but the plots show one is a straight line, another is curved, and others are shaped by outliers. 

**The problem:** If we only look at numbers, we may miss these hidden patterns and make wrong conclusions. 

**The solution:** Always graph the data before and after running models. Simple plots like scatterplots can reveal issues like outliers, non-linear trends, or unusual points.

### Loading and Exploring the Data

```{r}
#| label: load-data

# Load Anscombe's data
data(anscombe)
summary(anscombe)
```

### Building Linear Models

```{r}
#| label: build-models

# Create four model objects
lm1 <- lm(y1 ~ x1, data=anscombe)
lm2 <- lm(y2 ~ x2, data=anscombe)
lm3 <- lm(y3 ~ x3, data=anscombe)
lm4 <- lm(y4 ~ x4, data=anscombe)

# Compare model coefficients
ff <- y ~ x
mods <- setNames(as.list(1:4), paste0("lm", 1:4))

for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  mods[[i]] <- lmi <- lm(ff, data = anscombe)
}

# Show coefficients are identical
sapply(mods, coef)
```

### Visualizing the Four Datasets

```{r}
#| label: anscombe-plots
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Anscombe's Quartet: Four datasets with identical summary statistics but very different patterns"

# Set up plotting layout
op <- par(mfrow = c(2, 2), mar = 0.1+c(4,4,1,1), oma = c(0, 0, 2, 0))

# Plot all four datasets with regression lines
for(i in 1:4) {
  ff[2:3] <- lapply(paste0(c("y","x"), i), as.name)
  plot(ff, data = anscombe, col = "red", pch = 21, bg = "orange", cex = 1.2,
       xlim = c(3, 19), ylim = c(3, 13),
       main = paste("Dataset", i))
  abline(mods[[i]], col = "blue", lwd = 2)
}
mtext("Anscombe's 4 Regression Data Sets", outer = TRUE, cex = 1.5)
par(op)
```

### Key Insights

The visualization reveals four very different relationships:

- **Dataset 1:** Perfect linear relationship
- **Dataset 2:** Clear non-linear (curved) relationship  
- **Dataset 3:** Linear relationship with one outlier
- **Dataset 4:** No relationship except for one influential point

Despite these dramatic differences, all four datasets have:
- Same mean for x and yy

- Same correlation coefficient
- Same regression line
- Same R-squared value

This demonstrates why **visualization is essential** for understanding data patterns that summary statistics alone cannot reveal.

## 2. Changing Color in Fall Script

*[This section will be completed in the next iteration]*

## 3. Critique of Plot Design (Nathan Yau Principles)

*[This section will be completed in the next iteration]*